---
title: "Genralised Linear Models"
author: "Iliana Vasiliki Ntinou"
date: "6/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Exercises GLM

##Poisson GLM
1. Import sorich.csv
```{r}
rich <- read.csv("sorich.csv")
str(rich)
ggplot(rich, aes(x = cover, y = nsp)) + 
  geom_point() + 
  labs(y = "Species Richness", x = "Cover")
```

2. Fit a linear model for species richness (nsp) against cover assuming a Normal distribution
```{r}
fit_glm <- glm(nsp ~ cover, data = rich)
anova(fit_glm)
summary(fit_glm)

```

3. Study the diagnostic plots
```{r}
par(mfrow=c(2,2)) 
plot(fit_glm)
```
4. What distribution is the response?
*poison!*
5. Do an appropriate analysis
```{r}
fit1_glm <- glm(nsp ~ cover, data = rich, family = poisson)
anova(fit1_glm, test = "Chisq")
summary(fit1_glm)
ggplot(cbind(rich, fit = fitted(fit1_glm)), aes(x = cover, y = nsp)) + 
  geom_point() + 
  geom_line(aes(y = fit), colour = "red") +
  labs(y = "Species Richness", x = "Cover")
```

6. Check for over-dispersion. If necessary modify the model.
```{r}
anova(fit1_glm)
fit2_glm <- glm(nsp ~ cover, data = rich, family = quasipoisson) 
anova(fit2_glm)
```
7. Interpret the results
They are exactly the same.
8. How does the width of the confidence interval at cover = 10 change when over-dispersion is allowed for?

9. Do the grasses differ from the herb, i.e. make grassherb a factor and include it in the model and test its significance?


## How much does over-dispersion affect results
1. Use `sample()` to randomise the response in sorich to remove the relationship between the predictor and response.
2. Test if a Poisson GLM with cover as a predictor is significant
3. Repeat 1000 times and find the distribution of p-values

## Binomial GLM and GAM
1. Open library mgcv
2. Import data pot.csv
3. What type of distribution is the response variable?
4. What type of link-function do we use?
5. Do an appropriate GLM analysis?
6. Interpret the results?
7. Do a GAM analysis
8. Compare the GLM and GAM models.
9. Which model would you prefer, and why?