---
title: "Practical in model selection"
author: "Iliana Vasiliki Ntinou"
date: "6/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## hypothesis testing

### Import the palmerpenguin data and load libraries
```{r, message=FALSE}
#instal the package with: install.packages("palmerpenguins")
data("penguins", package="palmerpenguins")
penguins
#drop the NA values
penguins <- na.omit(penguins)
#libraries
library(tidyverse)
library(nlme)
```


### Test the hypothesis that bill length differs between species.

```{r}
lm1 <- lm(bill_length_mm ~ species, data = penguins)
anova(lm1)
penguins %>% 
  ggplot(aes(x = species, y = bill_length_mm, fill = sex)) +
  geom_boxplot()
```


### Test the hypothesis that bill length differs by sex in addition to species.How should the p-values be interpreted?

```{r}
lm1 <- lm(bill_length_mm ~ species, data = penguins)
lm2 <- lm(bill_length_mm ~ species + sex, data = penguins)
anova(lm1, lm2)
```
So we conclude from the p value that the second mdel with the sex is better. So bill length is also affected by the sex.

## Exploratory model building 

Normally you would be doing this on a separate data set

### Use forward selection to find the best model to explain bill length.

```{r, message=FALSE}
library(MASS)
mod <- lm(bill_length_mm ~.^2, data=penguins)
summary(mod)
model <- stepAIC(mod, k=2)
```
The model with the least AIC is 

bill_length_mm ~ species + island + bill_depth_mm + flipper_length_mm + 
    body_mass_g + sex + year + species:sex + island:bill_depth_mm + 
    bill_depth_mm:sex + bill_depth_mm:year + flipper_length_mm:body_mass_g + 
    flipper_length_mm:sex
    
    
### Build a set of candidate models to explain bill depth using one or two predictors.  

```{r}
mods <- list()
mods$m0 <- lm(bill_length_mm ~ 1, penguins)
mods$m1 <- lm(bill_length_mm ~ species, penguins)
mods$m2 <- lm(bill_length_mm ~ sex, penguins)
mods$m3 <- lm(bill_length_mm ~ sex + species, penguins)
mods$m4 <- lm(bill_length_mm ~ sex * species, penguins)

```


### Extract the AIC from each  models (hint use function `AIC`). Which is the better model?
```{r}
sapply(mods, AIC)

# OR
m0 <- lm(bill_length_mm ~ 1, penguins)
m1 <- lm(bill_length_mm ~ species, penguins)
m2 <- lm(bill_length_mm ~ sex, penguins)
anova(m0,m1,m2)
#model m1 has significant reduction of RSS
m3 <- lm(bill_length_mm ~ sex + species, penguins)
m4 <- lm(bill_length_mm ~ sex * species, penguins)
anova(m3,m4)
#the interaction term in m4 is not significant so we chosse m3
```


### Calculate the deltaAIC for each model and the AIC weights for each model. Interpret these weights.
ΔAIC - difference between the best AIC and AIC of the other models. Best model has a ΔAIC of zero, highest values for worst models

AIC weights exp(−0.5∗ΔAIC)∑exp(−0.5∗ΔAIC)
AIC weights – probability model is best of those tested
```{r}
library("AICcmodavg")
aictab(mods)
```


## Collinearity 

### Make a model predicting bill_length from all other variables. Find the VIF of each predictor. Are there any problem variables? 
```{r}
model1 <- lm(bill_length_mm ~., data = penguins)
library(olsrr)
olsrr::ols_vif_tol(model1)
```
Variance Inflation Factor
Measure of how much variance of the regression coefficient is “inflated” by correlation among the predictor variables

*1 is no correlation.*

*Values above 10 maybe a cause for concern.*

*Some argue for a VIF threshold of 3.*

### Use `GGally::ggpairs()` to plot the data to try to identify the cause of any high vif.


```{r, message=FALSE}
model1 <- lm(bill_length_mm ~., data = penguins)
library(GGally)
GGally::ggpairs(penguins)
``` 

### Use `MASS::mvrnorm()` to simulate 100 observation of two predictor variables (x and z) with a given correlation. Simulate a response variable y = b0 + b1x + b2z. Test how the uncertainty in the coefficients changes with the correlation (and hence vif) of the predictor variables.

```{r}
# data info

Sigma = matrix(c(1, 0.9, 0.9, 1), 2, 2)
Sigma
a <- mvrnorm(n = 100, mu=c(1,1), Sigma) %>% as.data.frame() %>% mutate(y= V1 +V2 +rnorm(100))
a
plot(a)
moda <- lm(y~., data=a)
olsrr::ols_vif_tol(moda)


```

