---
title: "Practical on GLS"
author: "Iliana Vasiliki Ntinou"
date: "6/2/2021"
output: html_document
---
## Load libraries

```{r, message=FALSE }
library(nlme) 
library(tidyverse)
```

Example using the built-in `iris` data.

```{r}
data(iris)
summary(iris)
```


Make a plot to show how `Petal.Length` varies between species.

```{r}
ggplot(iris, aes(x =Species, y = Petal.Length)) +
  geom_boxplot()
```


Find the variance of `Petal.Length` for each species.

```{r}
iris%>% group_by(Species) %>% 
  summarise(var = var(Petal.Length))
```


Fit an anova using `lm` between `Petal.Length` and species and examine the diagnostic plots.

```{r}
# the model
fit_lm <- lm(Petal.Length ~ Species, data = iris)
summary(fit_lm)

#diagnostic plots
par(mfrow=c(2,2))
plot(fit_lm)
```


Fit a `gls` for the same model. Have the coefficients changed?

```{r}
fit_gls <- gls(Petal.Length~Species, data=iris)
summary(fit_gls)
```

Fit a `gls` for the same model but allow the variance to be different for each species by adding `varIdent` to the `weights` argument of the model. 

```{r}
fit_gls1 <- gls(Petal.Length~Species, data = iris, weights = varIdent(form = ~ +1|Species))
summary(fit_gls1)
```

Use `AIC` to test if this is a better model.

```{r}
anova(fit_gls, fit_gls1)
```

