---
title: "Power Tests"
author: "Iliana Vasiliki Ntinou"
date: "5/28/2021"
output: html_document
runtime: shiny
---

### Load libraries

```{r,message=FALSE}
library(tidyverse)
library(lme4)
library(pwr)
```

### Run `biostats.apps::power_lm_app()` and explore the effect of the number of observations on the width of the confidence intervals, the proportion of trials that are statistically significant, and the effect size of significant trials.
```{r, message=FALSE}
biostats.apps::power_lm_app()
```

### Simulating a power test

A scientist is setting up an experiment to test whether a new type of food gives more weight gain in fish. They need to know how many fish they need to sample to have an 80% probability (i.e. power = 0.8) of getting a statistically significant result.

The experiment has a control (standard food) and a treatment arm (new food).

Fish typically have a mass of 1kg with a standard deviation of 0.2kg. The scientist hopes that the new food will give at least 0.1kg extra mass (otherwise it is not worth the extra cost).

*Simulate data for a single experiment with 10 fish in each arm of the experiment and test if your experiment was significant.*
```{r}
# data info
n <- 50 #number observations in each group
delta <- 0.1 #difference between means
sd <- 0.2 # standard deviation

# simulate means
mu <- rep(c(0, delta), each = n)

# add noise
y <- mu + rnorm(length(mu), sd = sd)

## predictor
x <- factor(rep(c("A", "B"), each = n))

# run test
test <- t.test(y ~ x)

test
broom::glance(test)
```

*Make your simulation into a function and repeat it 100 times. What proportion of trials are statistically significant?*
```{r}
sim_t_test <- function(n, delta, sd, ...){

mu <- rep(c(0, delta), each = n) # simulate means

y <- mu + rnorm(length(mu), sd = sd) # add noise

x <- factor(rep(c("A", "B"), each = n)) # predictor
  
test <- t.test(y ~ x) # run test

broom::tidy(test) %>% mutate(n = n, delta = delta, sd = sd)
}

sim_t_test(n = 50, delta = 0.1, sd = 0.2)


nrep = 100
control <- crossing(rep_no = 1:nrep, n = seq(10, 100, 20)) 

runs <- control %>% pmap_df(sim_t_test, delta = 0.1, sd = 0.2) %>% mutate(sig = p.value <= 0.05)
runs 

# see the proportion of trials that are significant
(sum ( runs$sig == TRUE )) / 500
```
*Plot your results.*
```{r}
p <- runs  %>%
  group_by(n) %>%
  summarise(power = mean(sig)) %>%
  ggplot(aes(x = n, y = power)) +
  geom_line() +
  geom_point()
p
```

*Run the simulation for different numbers of fish. How many are needed to get enough power?*
```{r}

```

### A second variable

The scientist thinks that a second variable, temperature, is also important, so expands the experiment to include this second variable with two levels, cool and warm.

*Extend your power analysis to cope with the new experimental design.*
```{r}

```

*How many fish do they need to get enough power.* 
```{r}

```

*Plot your result.*
```{r}

```


### Distribution of p values under Null hypothesis

*Simulate some data where the null hypothesis is true (ie effect size is zero).*
```{r}

```

*Plot the distribution of p-values.*
```{r}

```

*Compare this to the distribution of p-values calculated above.*
```{r}

```

