---
title: "Spatial & Temporal structure"
author: "Iliana Vasiliki Ntinou"
date: "6/2/2021"
output: html_document
---

## Simulation
1) Simulate a 100-observation autocorrelated timeseries with `arima.sim`, with a first order autoregressive coefficient of 0.5. Also make a time vector of `1:100`

```{r, message=FALSE }
#Load libraries
library(nlme) 
library(tidyverse)
library(lmtest)
library(broom.mixed)
```
```{r}
ts.sim<-arima.sim(list(order=c(1,0,0)#specifying that this is an AR1 model 1st order
                       , ar=0.5),#autoregressive coefficient
                  n=100) #number of obs
time <- c(1:100)
```

2) Plot the data. 

```{r}
ts.plot(ts.sim)
```


3) Regress the timeseries against time with an OLS model. Does the model appear to be statistically significant?

```{r}
lm.0 <- lm(ts.sim~time(ts.sim))
lm.0
anova(lm.0)

```


4) Plot the model diagnostics, including an `acf` and `pacf` of the residuals.
```{r}
acf(resid(lm.0))
pacf(resid(lm.0))
```

5) Use the Durbin-Watson test to test the residuals for autocorrelation.
```{r}
dwtest(lm.0)
```

6) Fit a gls with an appropriate correlation structure. Is this a better model? How have the p-value and effect size changed?
```{r}
gls.0 <- gls(ts.sim~time(ts.sim))
summary(gls.0)
gls.00 <- gls(ts.sim~time(ts.sim), corr=corAR1())
summary(gls.00)
anova(gls.0, gls.00)

```

7) Repeat the above 1000 times and find how autocorrelation affects the distribution of the p-value and the effect size.

```{r}
rerun(1000, {ts.sim<-arima.sim(list(order=c(1,0,0), ar=0.5), n=100) 
time <- c(1:100)
lm.0 <- lm(ts.sim~time(ts.sim))})%>% 
map_dfr(glance) %>% 
  ggplot(aes(x = p.value)) +
  geom_histogram()

```

## Real data

1) The built-in dataset LakeHuron has annual lake level data from 1885 to 1972
Load the data with the command data(LakeHuron)
```{r}
data(LakeHuron)
summary(LakeHuron)
```

2) Plot the data.
```{r}
plot(LakeHuron)
```

3) Regress the LakeHuron lake level against year using a linear model. Is there a significant trend?
```{r}
mod1<-lm(LakeHuron~time(LakeHuron))
summary(mod1)
```

4) Plot the autocorrelation and partial autocorrelation functions for the residuals from this regression. Interpret them.
```{r}
acf(resid(mod1))
pacf(resid(mod1))
```

5) Fit an autoregressive models to the residuals. Compare the results with your interpretation of the PACF plot.
```{r}
year<-time(LakeHuron)
fit.gls<-gls(LakeHuron~year)
summary(fit.gls)
```

6) Fit a gls model using a corAR1 correlation structure. Test if the correlation structure is necessary. Is the trend significant? 
```{r}
fit2.gls<-gls(LakeHuron~year, corr=corAR1())
summary(fit2.gls)
anova(fit.gls,fit2.gls)
```

7) Fit a gls model using a corARMA correlation structure with two AR terms. Is this model an improvement?
```{r}
fit3.gls <- gls(LakeHuron~year, corr=corARMA(p=2,q=0))
summary(fit3.gls)
anova(fit.gls,fit2.gls, fit3.gls)
```
