---
title: "LINEAR MODELS"
author: "Iliana Vasiliki Ntinou"
date: "5/28/2021"
output: html_document
---


### Load libraries

```{r,message=FALSE}
library(tidyverse)
library(lme4)
```

### Install and load the palmerpenguins package and see the penguins data

```{r}
#instal the package with: install.packages("palmerpenguins")
data("penguins", package="palmerpenguins")
penguins
#drop the NA values
penguins <- na.omit(penguins)
```

###  Do big birds have big beaks?
Fit a linear model between bill length and body mass, and interpret the coefficients.
```{r}
fit.lm<- lm(bill_length_mm~body_mass_g, data=penguins)
anova(fit.lm)
summary(fit.lm)
plot(penguins$body_mass_g, penguins$bill_length_mm)
lines(penguins$body_mass_g, fitted(fit.lm), col="red")
```
The coefficients can be written in a mathematical equation as:
y=(2.690e+01) + (4.051e-03)x 
2.690e+01 is the mean of the intercept and 4.051e-03 is the slope.

### Examine the diagnostics plots. Is everything OK?
Diagnostic plots focus on residuals of the model.
There are 4 questions we shoud be asking ourselves to see if everything is ok by looking at the diagmnostics plots
*1) Residuals vs Fitted values : Non-linear patterns?* 
The red line and points should not show a clear pattern.
*2) Normal Q-Q plot: Normal distribution of residuals?*
The data points should follow the straight dashed line.
There will always be some deviations but they should not deviate severely.
*3) Scale Location: Constant variance?*
You should see a fairly horizontal line with randomly spread points.
*4) Residuals vs Leverage: Influential outliers?*
If you have points placed outside the dashed lines, they have high Cook's distance. 
This means that they are influential outliers.
Removing them will change the results of the regression.
```{r}
par(mfrow=c(2,2))
plot(fit.lm)
```
So the answer in this case is that everything looks ok!

### Calculate the residual sum of squares (find sum of the squared difference between the estimates calculated from the coefficients and the observed values).
The mathematical explanation would be:
The model is yi = β0 + β1xi + εi 
       where εi = error (residuals= actual - predicted)
●*Σεi^2*= Σ(yi - (β0 + β1xi))^2
       where β1 = sxy / sx^2 = covar(xy)/var(x), 
             β0 = mean(y) - β1mean(x)
A fast way to do it is by looking the output of ANOVA
```{r}
#in this case  it is y=(2.690e+01) + (4.051e-03)x
sum((penguins$bill_length_mm - (penguins$body_mass_g*0.004051 +26.9))^2)


#yi = β0 : The null model is basically true when the null hypothesis is true
fit0.lm <-lm(bill_length_mm~+1, data=penguins)
anova(fit0.lm)
#the residual sum of squares predicted by the null model is 9928.9
#yi = β0 + β1xi :
fit.lm <- lm(bill_length_mm~body_mass_g, data=penguins)
anova(fit.lm)
#the residual sum of squares predicted by our model is 6479.1
anova(fit0.lm, fit.lm, test="F") # to compare both RSS values

#The RSS is significantly reduced by the model compared to the null model
```

If we recalculate the Residual Sum of Squares (RSS) using different values for the coefficients do we observe if the new RSS is larger or smaller?
```{r}

```

### When we estimate the slope using covariance and variance we can examine if we get the same result?

As we explained before: β1 = sxy / sx^2 = covar(xy)/var(x)
```{r}
sxy <- cov(penguins$body_mass_g, penguins$bill_length_mm)
sx2 <- var(penguins$body_mass_g)
slope <- sxy/sx2 
slope
```
The intercept from the model is 4.051e-03 which is almost equal to the result we calculated

### Adding species to the model as a predictor, does it improve the model and how should the coefficients be interpreted?
 General interpretation: The changes in bill length with changes in bodymass depends on the species, or vice versa.
```{r}
#In that case the interpretation of the model would be: The slope of the continuous predictor (body mass), depends on the levels of the categorical predictor (species).
fits.lm <-lm(bill_length_mm~body_mass_g*species, data=penguins) 
anova(fits.lm) 

# the interaction term is not significant (p-0.1826) so we remove it following the backward elimination technique

fits1.lm <- lm(bill_length_mm~body_mass_g + species, data=penguins)
anova(fits1.lm)
summary(fits1.lm)

anova(fits.lm, fits1.lm, test="F")

```
The coefficients from the summary output can be interpreted as:
Intercept: Mean bill length for Adelie species when bodymass=0.
body_mass_g: The slope of the line for Adelie species.
Tomatoresistant: The difference in intercept (mean at Nitrogen=0)
speciesChinstrap: The difference in intercept (mean at bodymass=0)for Chinstrap species compared to Adelie.
speciesGentoo: : The difference in intercept (mean at bodymass=0)for Gentoo species compared to Adelie.
