---
title: "PRACTICAL IN EXPLORATORY ANALYSIS"
author: "Iliana Vasiliki Ntinou"
date: "5/26/2021"
output: html_document
---

# Practice in exploratory analysis
## Load libraries

```{r,message=FALSE}
library(tidyverse)
```

## Install and load the palmerpenguins package and see the penguins data

```{r}
#instal the package with: install.packages("palmerpenguins")
data("penguins", package="palmerpenguins")
penguins
#drop the NA values
na.omit(penguins)
```

## Find the range, mean, median and variance of variable bill_length_mm

```{r}
#it will show, minimum, maximum, median and mean
summary(penguins$bill_length_mm, na.rm=TRUE)
#it will show variance
var(penguins$bill_length_mm, na.rm=TRUE)

```

## Find the standard error of the mean of bill_length_mm

```{r}
#install package plotrix if not installed: install.packages("plotrix")
#load library
library(plotrix)
std.error(penguins$bill_length_mm, na.rm=TRUE)
```

## Plot a histogram of the variable bill_length_mm showing mean and median

```{r}
p1 <- ggplot(penguins, aes(bill_length_mm)) + 
      geom_histogram(col="black", fill="green")+  #modify the colors
#add mean      
      geom_vline(aes(xintercept = mean(bill_length_mm, na.rm=TRUE), ), color="blue") + 
#add median      
      geom_vline(aes(xintercept = median(bill_length_mm, na.rm=TRUE), ), color="red")+ 
      theme_bw()+  #choose a friendly theme
      labs(title="Histogram for Bill length", x="Bill length in mm", y="Frequency") #add labels
p1
```

## Plot bill_length_mm against bill_depth_mm

```{r}
p2 <- ggplot(penguins, aes(x=bill_length_mm, y=bill_depth_mm ))+
      geom_jitter()+
      theme_bw()+
      labs(x="Bill length in mm", y="Bill depth in mm") 
p2
```

## Find the covariance and correlation of bill_length_mm against bill_depth_mm

```{r}
#covariance
cov(penguins$bill_length_mm, penguins$bill_depth_mm, use="pairwise.complete.obs")
#correlation
cor(penguins$bill_length_mm, penguins$bill_depth_mm, use="pairwise.complete.obs")
#test if the correlation is significant
cor.test(x=penguins$bill_length_mm, y=penguins$bill_depth_mm, use="pairwise.complete.obs")
```


## Plot bill_length_mm against bill_depth_mm, coloured by species

```{r}
p3 <- ggplot(penguins, aes(x=bill_length_mm, y=bill_depth_mm, colour=species))+
      geom_jitter()+
      theme_bw()+
      labs(x="Bill length in mm", y="Bill depth in mm")
p3
```

## Find out if the correlation between bill_length_mm against bill_depth_mm is significant for any of the species

```{r}
penguinsA <- filter(penguins, species=="Adelie")
cor.test(x=penguinsA$bill_length_mm, y=penguinsA$bill_depth_mm, use="pairwise.complete.obs")
penguinsC <- filter(penguins, species=="Chinstrap")
cor.test(x=penguinsC$bill_length_mm, y=penguinsC$bill_depth_mm, use="pairwise.complete.obs")
penguinsG <- filter(penguins, species=="Gentoo")
cor.test(x=penguinsG$bill_length_mm, y=penguinsG$bill_depth_mm, use="pairwise.complete.obs")
```

This is a *subgroup analysis* which includes some **dangers**... 
One of the problems with subgroup analysis is that it is highly likely to get a statistically significant false result, meaning the more groups one includes in their analysis the higher the chance to find a statistically significant effect. 
Sometimes it all comes down to the group size that is being investigated, meaning that there may be simply needed a larger group size to detect an effect if there is one.
 
# Extra Practice

## Install the datasauRus package and load it with library.

```{r, message=FALSE}
#instal the package with: install.packages("datasauRus")
library(datasauRus)
```

## Dataset datasaurus_dozen has x and y variables and a dataset name.

```{r, }
data("datasaurus_dozen", package="datasauRus")
datasaurus_dozen
#simple plot
pdd<- ggplot(datasaurus_dozen, aes(x=x, y=y, colour=dataset))+
      geom_point()+
     facet_wrap(~dataset)
pdd
```

## Calculate the mean and standard deviation of x and y and the correlationn between x and y, for each dataset.

```{r}
datasaurus_dozen %>% group_by(dataset) %>% 
                     summarise(meanx = mean(x), standarddeviationx = sd(x),
                               meany = mean(y), standarddeviationy = sd(y),
                               COR=cor(x,y))
```
